Epistemic Semantics in Guarded String Models

### 2021-01-12 tex/pdraft.tex
Initiate publication draft pdraft.tex
Add page 9, add appendix, change to nop events in Figure 2.

%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

%\author{Eric Campbell \\
%  Cornell University\\
%  \texttt{ehc86@cornell.edu} \\\And
%  Mats Rooth \\
%  Cornell University\\
%  \texttt{mr249@cornell.edu} \\}

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents
semi-privately looking at the coin, and events of secretly turning the coin.
The line beginning with {\tt
  state} enumerates ${\sf{B}}$. The line beginning with {\tt restrict}
gives the state formula.  The lines beginning with {\tt event} declare
events and their effect formulas.  Finally the lines beginning with
{\tt agent} define {\em event alternative} relations for agents.  Each
clause with an arrow has a single event symbol on the left, and a
disjunction of alternative events on the right of the arrow. The
interpretation of Amy's alternatives for $b_1$ (Bob peeks at heads),
is that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bob peeks
at tails) could be happening. Her alternatives for $a_{01}$ and
$a_{10}$ (she turns the coin over) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. Similarly, Bob
secretly turns the coin over, in $b_{10}$ or $b_{01}$, she doesn't
know anything has happened, so her alternatives are the no-operation
events $o_1$ and $o_0$ for heads-worlds and tails-worlds
respectively. Bob's alternatives are the same, \emph{mutatis mutandi}.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epik program describing a possible-worlds event sequence
  model for two agents with information about one coin, and events of
  the agents semi-privately looking at the coin. }
\label{fig:epik-example}
\label{epikfigure}
\end{figure}

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
  \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a logical language that extends the Epik term language with lambda. }
  \label{lexiconfigure}
\end{figure}

@@

@@

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by the grammar.
Logical relations between propositions are checked in the finite state calculus by
checking set-theoretic relations between sets of worlds.
For instance entailment $p \rightarrow q$ is decided by checking in an interpreter
for the finite state calculus whether $p - q$ is non-empty.
In the model defined by Figure 2, the
propositions \exi{three}{b} and \exi{three}{e} are
independent. They are equivalent in a version 
without secret flipping.

@@

\section{Discussion}
The methodology presented here is designed for use in research in linguistic
semantics, and for education at the level of a second graduate course in formal
semantics, covering intensionality.  There are straightforward extensions to
additional linguistic phenomena, such as tense and perfective aspect as in
\exi{extension}{a}, and the combination of metaphysical modality and prospective
aspect in \exi{extension}{b}.

@@

The development here is concerned with defining concrete computable possible
worlds models, and applying them in natural language semantics.  
Issues for further investigation are mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms, coalgebra, and decidability.

We will release Epik's source code under an open source license prior
to the conference.

\bibliographystyle{acl_natbib}
% \bibliographystyle{chicago}
\bibliography{fs}

>>>>

% MR249 UNCOMMENTED FOR FINAL VERSION
\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\author{Eric Campbell \\
  Cornell University\\
  \texttt{ehc86@cornell.edu} \\\And
  Mats Rooth \\
  Cornell University\\
  \texttt{mr249@cornell.edu} \\}

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.
Finally the lines beginning with {\tt agent} define {\em event
  alternative} relations for agents.  Each clause with an arrow has a
single event symbol on the left, and a disjunction of alternative
events on the right of the arrow. The interpretation of Amy's
alternatives for $b_1$ (Bob peeks at heads), is that when $b_1$
happens, for Amy either $b_1$ or $b_0$ (Bob peeks at tails) could be
happening. Her alternatives for $a_{01}$ and $a_{10}$ (she turns the
coin over in one state or the other) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. When Bob secretly
turns the coin over, in event $b_{10}$ or $b_{01}$, she doesn't know
anything has happened, so her alternative is the no-operation,
no-information event event $o$. Bob's event relation is symmetric.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o  h:h + t:t\\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o  -> o \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epik program describing a possible-worlds event sequence
  model for two agents with information about one coin, and events of
  the agents semi-privately looking at the coin, and privately turning
  the coin.}
\label{fig:epik-example}
\label{epikfigure}
\end{figure}

@@

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
  \end{tabular}
    \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a language that extends the Epik logical language with lambda. The complete grammar
    is in the appendix.}
  \label{lexiconfigure}
\end{figure}

@@

\end{examples}

Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by
the grammar.  Logical relations between propositions are checked in
the finite state calculus by checking set-theoretic relations between
sets of worlds.  For instance entailment $p \rightarrow q$ is decided
by checking in an interpreter for the finite state calculus whether $p
- q$ is non-empty.  In the model defined by Figure 2, the propositions
\exi{three}{b} and \exi{three}{e} are independent. They are equivalent
in a version without secret flipping.

@@

\section{Discussion}
The methodology presented here is designed for use in research in linguistic
semantics, for research in computational syntax and semantics on model-theoretically
grounded semantics, 
and for education at the level of a second graduate course in formal
semantics, covering intensionality.  There are straightforward extensions to
additional linguistic phenomena, such as tense and perfective aspect as in
\exi{extension}{a}, and the combination of metaphysical modality and prospective
aspect in \exi{extension}{b}.

@@

The syntactic part of the grammar formalism uses categorial grammar to
map between strings (or trees or derivations) and logical
translations.  This way of using a categorial grammar is standard
\citep{steedman2000syntactic,bozsahin2012combinatory}, and there are
computational implementations for it
\citep{barker2005parser,bozsahin2021ccglab}.\footnote{Our grammar uses
  basic categorial grammar, not additional features such as
  combinators, type raising, continuations, or multimodal slashes that
  interact with the grammar in non-trivial ways.  These features would
  become relevant in more sophisticated grammars.  Multimodal
  categories are used for phrases such as a tenseless verb phrase that
  semantically are functional, but can combine in the grammar only as
  arguments.}  While this way of proceeding is simple and attractive
for the applications we are interested in, it would be possible to use
our semantics with other frameworks that have a semantics that uses a
lambda extension of a logical language.

%\newcommand{\Prime}[1]{{\rm\bf{#1}}}

The semantics presented here is exclusively concerned with events and
worlds.  Semantic type systems for natural language usually include a
type for individuals.  In the coin example, the individuals are hidden
in the primitive event symbols such as $a_1$ and $b_1$. This situation
would get worse in a more elaborate model with more agents and
multiple coins.  The solution to this should be to base the world
construction on grounded event terms such as
$\Prime{look}(\Prime{a},\Prime{k},1)$, for ``agent $\Prime{a}$ see
that coin $\Prime{k}$ is heads'', rather than atomic events.  This way
of proceeding is found in research on situation calculus
\citep{reiter2001knowledge}.  How to incorporate it in the scheme for
Epik specifications and into the computational parts of the proposal
is a topic for future research.  In this it is not immediately clear
whether it is possible to introduce quantification over individuals,
together with individual constants, while maintaining the finite state
semantics. \citet{rooth2017finite} develops an approach to a positive
answer, based on introducing markers for witnesses for discourse
referents in the construction of worlds.

The development here is concerned with defining concrete computable possible
worlds models, and applying them in natural language semantics.  
Issues for further investigation are mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms and coalgebra.

Source code, examples, and instructions for Epik are distributed at
\url{https://github.com/ericthewry/epik}.

\bibliographystyle{acl_natbib}
% \bibliographystyle{chicago}

\clearpage 
\bibliography{fs}

\clearpage

\onecolumn

\section*{Appendix}
Complete categorial grammar extending Figure \ref{lexiconfigure}.  The online supplement
includes a version of this grammar that can be run using Barker and Shan's
chart parser for categorial grammar \citep{barker2005parser}.

  \begin{tabular}[t]{lcc}
    \toprule
    \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule
    \multicolumn{3}{c}{{\bf Statives with expletive subject}} \\ 
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\ \hline
    \multicolumn{3}{c}{{\bf Agents}} \\ 
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\ \hline
    \multicolumn{3}{c}{{\bf Auxiliary verbs and expletive subjects}} \\        
        it & $d$ & {\em{dummy}} \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        it's & $(t \slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        isn't & $t\slash(d \backslash_D t)$ & $\lambda Px.\neg P x$ \\  
        doesn't & $(e \backslash t)\slash(d \backslash_V t)$ & $\lambda Px.\neg Px$ \\  \hline        
    \multicolumn{3}{c}{{\bf Tensed attitude verbs}} \\        
        knows & $(e \backslash t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline
    \multicolumn{3}{c}{{\bf Base form attitude verbs}} \\        
        know & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believe & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline       
    \multicolumn{3}{c}{{\bf Complementizers}} \\        
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
    \multicolumn{3}{c}{{\bf Complementizers for base form verbs}} \\        
        that  & $(((e \backslash_V t){\slash_M}t) \backslash (e \backslash_V t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash_V t){\slash_M} t) \backslash (e \backslash_V t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ % \bottomrule    
      \end{tabular}
\end{document}

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
    \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule
    \multicolumn{3}{c}{{\bf Statives with expletive subject}} \\ 
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\ \hline
    \multicolumn{3}{c}{{\bf Agents}} \\ 
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\ \hline
    \multicolumn{3}{c}{{\bf Auxiliary verbs and expletive subjects}} \\        
        it & $d$ & {\em{dummy}} \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        it's & $(t \slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        isn't & $t\slash(d \backslash_D t)$ & $\lambda Px.\neg P x$ \\  
        doesn't & $(e \backslash t)\slash(d \backslash_V t)$ & $\lambda Px.\neg Px$ \\  \hline        
    \multicolumn{3}{c}{{\bf Tensed attitude verbs}} \\        
        knows & $(e \backslash t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline
    \multicolumn{3}{c}{{\bf Base form attitude verbs}} \\        
        know & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believe & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline       
    \multicolumn{3}{c}{{\bf Complementizers}} \\        
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \hline
    \multicolumn{3}{c}{{\bf Complementizers for base form verbs}} \\        
        that  & $(((e \backslash_V t){\slash_M}t) \backslash (e \backslash_V t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash_V t){\slash_M} t) \backslash (e \backslash_V t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ % \bottomrule    
      \end{tabular}
    \caption{Categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a language that extends the Epik logical language with lambda.}
  \label{lexiconfigure}
\end{figure}

Old version of grammar figure.

  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
